\setchapterpreamble[ur][.6\textwidth]{%
\dictum[Miyamoto Musashi, \textit{A Book of Five Rings} (circa 1645)]{%
It is difficult to understand the universe if you only study one planet.}\vskip1em
\dictum[Albert Camus, \textit{La Peste} (1947)]{%
Le mal qui est dans le monde vient presque toujours de l'ignorance, et les bonnes intentions peuvent faire du mal autant que la malveillance si elles n'ont pas la compr\'ehension. / The evil that is in the world always comes of ignorance, and good intentions may do as much harm as malevolence, if they lack understanding.}\vskip1em
}

\chapter[\texorpdfstring{Chapter 6 \\ General discussion}{Chapter 6 General discussion}]{General discussion}
\chaptermark{General discussion}
\label{chap:discu}

\section{Overview}
In this thesis, I investigated the causes and consequences of variation in fitness in a wild snow vole population. I showed that in this population, variation in proxies of individual fitness is not purely stochastic, but is underlain by variation in latent fitness (chapter \ref{chap:dynhet}). Furthermore, this variation in latent fitness has an additive genetic component, revealing the existence of natural selection and adaptive evolution (chapter \ref{chap:stasis} and \ref{chap:flusel}).
I also explored ways to decompose the causes of phenotypic changes and identified the quantitative genetic animal model as a appropriate tool to estimate evolution (chapter \ref{chap:decpop}).
Using this tool in various ways, I showed that body mass is an important contributor of variation in fitness proxies (chapter \ref{chap:stasis}), but not in a consistent way over time (chapter \ref{chap:flusel}). Nevertheless, body mass was a consistent contributor to genetic variation for fitness (chapters \ref{chap:stasis} and \ref{chap:flusel}), and body mass therefore evolved over the study period.

Below, I will further discuss both the insights brought by this thesis and the remaining challenges when it comes to understanding the causes of phenotypic variation and how wild populations respond to environmental change. 

\section{The causes of phenotypic variation}
This thesis has brought some new insights into the causes of phenotypic variation in general, and fitness variation in particular, but there still is much more to learn. Below, I comment on two promising directions that were only touched upon in this thesis, but that have the potential to improve the predictive understanding of the causes and consequences of fitness variation. These are the effects an individual's genes may have on other individuals, and the study of the molecular basis of genetic variation using genomic methods.

\subsection{The effect of the genes of others}
Using quantitative genetic methods, I decomposed the phenotypic variation of morphological and life-history traits into components related to additive genetic effects, maternal effects and permanent environments. This decomposition was sufficient to measure the rate of evolution of the direct genetic effects of an individual's genes (as summarised by its breeding value) (chapter \ref{chap:stasis}), that is, the direct action of an individual's genes on itself.
However, an individual's genes may have effects reaching out beyond itself, to the environment, and thus other individuals \parencite{Dawkins1982}. This may happen through interactions between individuals (indirect genetic effects, e.g. maternal effects, \cite{McAdam2014}), or through the pleiotropic action of genes expressed in kin at different life-stages (e.g. genetic conflict, \cite{Trivers1974}).

Indirect genetic effects could play an important role in shaping selection and evolution in the snow vole population. Indeed, in the snow voles, genes within an individual are likely to affect the phenotype of another individual during at least two types of situations. First, related females tend to form clusters of territories, and the presence of kin could suppress reproduction in subordinate females \parencite{Garcia-Navas2016}. Moreover, as in all placental mammals, maternal effects on offspring phenotypes are prevalent from pregnancy to weaning. 
Maternal effects have been studied extensively in natural populations \parencite{Wolf2009}, but estimates of the genetic component of maternal effects remain scarce \parencite{McAdam2014}. Nevertheless, genetic maternal effects may add to any direct genetic variation and thereby increase evolutionary potential \parencite{Mcglothlin2014, McAdam2014, Mcfarlane2015}. In the snow vole population, preliminary analyses have shown the presence of additive genetic maternal effects for body mass (results not shown), and these could be subject to selection and evolve adaptively. Note that in chapter \ref{chap:stasis}, maternal genetic effects were not explicitly modelled, and their evolution was instead assigned to phenotypic plasticity. A full account of body mass evolution should explicitly capture this evolution in addition to that of direct additive genetic effects. 

In addition to indirect genetic effects, the effect of others' genes matters for evolution in the presence of genetic conflicts, that is, genetic trade-offs between traits expressed in different individuals.
For four decades, genetic conflicts between parents and offspring have been thought to be a major constraint on the evolution of size \parencite[since][]{Trivers1974}, but the idea resisted empirical tests despite behavioural studies showing patterns consistent with it \parencite{Kolliker2015}. \cite{Kolliker2015} demonstrated that a genetic trade-off between offspring number and offspring size constrains the evolution of size in earwigs (\textit{Forficula auricularia}, Linnaeus 1758). Moreover, \cite{Rollinson2015b} presented qualitative evidence suggesting that this constraint is widespread among animals and could be a general explanation for the evolutionary stasis of size. 
In chapter \ref{chap:stasis} we briefly explored the possibility that a genetic conflict constrains the evolution of body mass, and found qualitative evidence that this is not the case. The snow vole study system is not an ideal to test this hypothesis, however. First, we do not capture all juveniles\textemdash some die or emigrate before their first year\textemdash and we cannot measure litter size accurately. Because mass is under selection in juveniles, selective disappearance is likely to blur the trade-off signal \parencite{Hadfield2011}. Second, whereas the genetic size-number trade-off is an explanation for evolutionary stasis of body size or mass,  mass is evolving in the snow vole population (chapter \ref{chap:stasis}). This makes it more difficult to formulate an expectation for the genetic covariance between mass and litter size in the presence of such a trade-off. 
Finally, although in theory it is possible to measure the genetic trade-off using quantitative genetics, the exact model that would need to be fitted has as of yet not been described and the modelling tools that are required are currently unavailable \parencite{Hadfield2012, Rollinson2015b}. This means that at the moment an experimental approach remains the only option to quantitatively test for a size-number genetic trade-off \parencite{Kolliker2015}, and such an approach appears impossible in a wild population such as Churwalden's snow voles. 

\subsection{Molecular basis of genetic variation}
On several occasions during this PhD, we considered using high-throughput sequencing \parencite{VanDijk2014} to  retrospectively obtain high-density marker data for the snow vole population (tissue samples are kept in $-80^\circ$ freezers for most of the individuals trapped in the last ten years).
As of yet we did not obtain the funding necessary for this, and time-limitations prevented further lab work and the development of a bioinformatic pipeline. 
As I discussed in chapter \ref{chap:intro}, molecular approaches to measuring selection and evolution are in general inferior to quantitative genetic approaches. Nevertheless, individual-based genomic data could bring complementary insights to my empirical chapters.

To start, individual-based genomic data could (marginally) improve quantitative genetic parameter estimates \parencite{Berenos2014} by: (i) allowing for the use of \emph{realized} relatedness in quantitative genetic animal models, instead of the relatedness \emph{expected} from the pedigree; and (ii) providing some relatedness information about individuals with unknown parents (for which there is no information at all in a pedigree). 
More importantly, individual-based genomic data would allow the identification of some of the genetic loci underlying phenotypic variation and quantitative evolution.
This task is generally a challenging one in small populations \parencite{Wellenreuther2016}, but the snow vole population presents three rare advantages that would ease it considerably. 

First, at least one trait, body mass, has been evolving during the last decade, and some adaptive molecular evolution must have happened. The search for the molecular basis of evolution would therefore start with the knowledge that there is something to find, and with indications on what functional types of genes are likely to be involved. 
Second, because in natural populations with complex demographics and mating patterns, there is in general no null-expectation for the effect of drift. This makes it difficult to show that evolution at a genetic locus is due to selection and not only due to drift. A pedigree provides such a null expectation, however. Simulating the random dropping of alleles down our pedigree would results in a null distribution of changes in allele frequencies against which to test for the effect of selection on each genetic locus. This method was successfully employed to show contemporary adaptive evolution at 67 genetic loci in a wild population of Florida scrub-jays (Nancy Chen, Evolution conference, 2016, Austin, USA).
Third, thanks to the availability of life-history data, it would be possible to correlate the allelic variation of the evolving loci to success and failure in various life-stages.  
Therefore, the combination of genomic and life-history data can pinpoint when selection occurs in life, and what kind of molecular mechanism selection acts on.
Altogether, individual-based genomic data could therefore refine not only our molecular understanding of phenotypic variation, but also provide clues regarding the ecological nature of selection. 

%\subsection{Origin and maintenance of variation in fitness}
%%fondamental maintenance of g var
%Still, our study is a snapshot, too short to expect significant loss of genetic variation. What would happen if the observed selective pressures would carry on? 
%Fluctuating selection might contribute to the maintenance of genetic variation
%New mutations certainly contribute, but probably not sufficient in general (charlesworth).
%Population structure. 

\section{Predicting responses to environmental change}
Anthropogenic environmental change has stimulated research aiming at understanding and predicting the response of natural populations to environmental change \parencite{parmesan2006, Chevin2012, Smallegange2013, Charmantier2014climate}, but massive challenges continue to hinder this research agenda. At the moment, it is already the retrospective study of phenotypic and demographic responses that remains inconclusive \parencite{Merila2001,McCarty2001, Charmantier2014climate, Brookfield2016}, and in most cases prospective prediction therefore seems to be well out of reach. 
During my Ph.D., I confronted three challenges that must be tackled to improve the predictive abilities of evolutionary ecology. Below, I discuss the problems with measuring selection, predicting the response to selection, and integrating evolutionary and demographic responses.

\subsection{Measuring selection in the wild}
For over 150 years, natural selection has been known to cause the match between organisms and their environment, and consequently biologists have attempted to understand its causes and mechanisms. More recently, the study of selection assumed a more applied goal as researchers hope to predict the response of natural populations to the selective pressures imposed by environmental change \parencite{Chevin2010a, Coulson2010, Merila2014}.
The principle of natural selection appears so simple: In a given environment, individuals with a phenotype that favours survival and fertility contribute more to the next generation. Given the level of research attention devoted to such a simple process, it may at first sight seem surprising that our understanding of natural selection has developed so slowly, and that its study remains so difficult. 
For most of the 20th century, the main brake to progresses was the lack of an unified framework to quantify selection in natural populations \parencite{Wade2006}. Such a framework progressively emerged, starting with covariance-based methods \parencite{Robertson1966, Price1970} which efficiently measure the total effect of selection. The most influential breakdown was the popularization of regression-based methods \parencite{Lande1979,Lande1983} which measure the proportional effect of selection per unit of phenotypic variation, and allows decomposing selection into the direct and indirect effects of selection on multiple traits \parencite{Broodie1995}.
Since then, these methods have provided thousands of estimates of selection in natural populations \parencite{Kingsolver2001,Stinchcombe2008,Kingsolver2012}. This has revealed several general patterns. For instance, directional selection is stronger and more common than suggested by early evolutionists, whereas stabilizing selection appears to be rare. Also, fertility selection is generally stronger than viability selection \parencite{Kingsolver2012}.
The abundance of estimates of selection should not be mistaken for a good understanding of natural selection, however. The estimation of selection through regression-based methods faces at least three difficulties that might severely hamper their significance and explain the general absence of genetic responses to selection \parencite{Merila2001, Brookfield2016}.

First, to obtain an unbiased measure of selection, fitness should be regressed on the trait of interest. Since, fitness is rarely observable directly, fitness proxies must be used instead. Hence, many estimates of selection are computed on fitness components, for instance fertility and survival \parencite{Kingsolver2012}. In this case, the estimation of selection can be biased in the presence of a trade-off between fitness components: a certain phenotype might increase survival but decrease fertility, so that the net selection on the trait is null, despite covariation with both fitness components \parencite{Thompson2011, Kingsolver2012, Brookfield2016}. Fortunately, this bias in general appears to be minor, with the exception of body mass \parencite{Kingsolver2011}.
For the empirical part of this thesis (chapter \ref{chap:stasis} and \ref{chap:flusel}), I used fitness proxies that attempted to include all fitness components in order to avoid such biases. Thus, I used lifetime reproductive success when measuring selection within a generation, and annual reproductive success plus twice survival when measuring selection within a year. These fitness proxies are imperfect since we do not capture all juveniles and a trade-off between early juvenile survival and reproduction could bias the selection estimation \parencite{Hadfield2008}. Still, estimates of evolution using the Price equation (that is, selection on the genotype) or using the trend in BLUPs for breeding values (that is, not using any information about selection nor fitness) agree qualitatively with my corrected estimates of selection (chapter \ref{chap:stasis} and \ref{chap:flusel}), suggesting that my proxies for fitness are adequate. 

Second, it is possible to estimate the total effect of selection on a trait with selection differentials, but it is much more difficult to disentangle the causal selective effect of a trait from the indirect selection due to other traits. In theory, it is possible to disentangle direct and indirect selection by including all the traits under selection in the analysis \parencite{Lande1983}. In natural populations, however, it is impossible to know \emph{a priori} what traits are under selection, and often it is impossible to measure all relevant traits \parencite{Brookfield2016, Hadfield2008}. Furthermore, as more traits are included in a selection analysis, the statistical power to detect significant selection on any one trait decreases \parencite{Mitchell-Olds1987}.
I did detect significant indirect selection on body mass, but genetic correlations between the traits considered were such that the prediction of evolution was not affected by the inclusion of indirect selection (chapter \ref{chap:stasis}). Only three traits were tested, however, and we cannot exclude that body mass is not under indirect selection. The evolution of body mass could be driven by selection on an unmeasured trait.
Nevertheless, this problem is irrelevant to the measures of total selection and evolution, on which chapters \ref{chap:stasis} and \ref{chap:flusel} rely.

Third, using covariance- and regression-based methods to estimate phenotypic selection essentially measure the statistical association between traits and relative fitness. Selection is however a causal association, be it direct or indirect. If the statistical association is entirely mediated by an environmental covariance between traits and fitness, there is no selection and no possibility of a genetic response to selection \parencite{Price1989, Rausher1992}. 
Body mass, the main trait analysed in this thesis, is likely to be very sensitive to this source of bias. Indeed, a favourable environment\textemdash for instance food-rich and parasite-free\textemdash is likely to lead to large mass, high survival, and high fertility. Accordingly, phenotypic estimates of natural selection on mass and size are overwhelmingly positive, but mass does not evolve as predicted from its heritability \parencite{Blanckenhorn2000, Kingsolver2012}. In the snow voles, an excess of environmental covariance does underlie the apparent selection on mass (chapter \ref{chap:stasis}).
A solution to the problem is the experimental manipulation of the trait of interest. This can break the link between phenotype and individual quality and reveals the causal action of phenotype on fitness components \parencite[e.g.][]{Tinbergen2004, Tschirren2006}. Still, experimental manipulation is not without its own limitations.
First, manipulations are work-intensive, time-consuming and must be designed carefully in order to manipulate the trait of interest without affecting any other traits. Moreover, manipulations cannot easily be applied to all traits. While an experimental approach has been widely used to study selection on brood size, it is not clear to me how one could manipulate body mass in a controlled way (that is, without accidentally affecting other traits).
Instead of an experiment, my approach to the challenge of environmental covariation has been to use quantitative genetics to identify the target of natural selection (chapter \ref{chap:stasis}). After having shown on-going adaptive evolution, I decomposed phenotypic selection into an additive genetic and an environmental component, for various fitness components. I found that only juvenile viability selection showed an additive genetic component, and according to the Robertson-Price identity, was the source of adaptive evolution. This made understanding the mechanism of this selection and measuring its strength a matter of hypothesis testing.
This approach could be used in other systems, but it requires sufficient phenotypic and relatedness data to fit bivariate animal models. In addition, although in the snow vole a single fitness component drove evolution, multiple fitness components could be involved, which would complicate the analysis. Finally, identifying the right fitness component(s) does not guarantee that the phenotypic mechanism of selection can be identified. A good understanding of the biological system will be necessary to formulate a reasonable hypothesis for the cause of selection. The testability of this hypothesis will also depend on data availability and quality, and will be subject to the limits of hypothesis testing approaches: there is always a risk of false-positives and a correlation does not prove causation.


\subsection{Evolutionary response}
Once a measure of phenotypic selection and a heritability estimate is obtained, it is straightforward to formulate a prediction of genetic response based on the breeder's equation \parencite{Lush1937, Falconer1996}. We have already seen (chapter \ref{chap:stasis} and \ref{chap:flusel}) that such a prediction is often unreliable in natural populations, however.  Estimates of selection might not correspond to causal selection, and unmeasured selection acting on genetically correlated traits might constrain evolution. I have shown that estimating the genetic component of selection, or the rate of evolution, can test whether selection has been measured appropriately to be predictive (chapter \ref{chap:stasis}). 

Nevertheless, most attempts to understand the evolutionary response to environmental change do not measure genetic parameters. Thus, the alarming lack of evidence for evolutionary responses to climate change probably originates primarily from a lack of tests for genetic change \parencite{Charmantier2014climate,Gienapp2014,Merila2014,Crozier2014}.
Ignoring the genetic properties (e.g. the heritability) of the trait of interest \parencite[e.g.][]{Forcada2014, Coulson2014, traill2014demography} easily leads to underestimating, or incorrectly dismissing, the potential to respond to selection and the actual evolutionary response \parencite{Nietlisbach2015, Chevin2015a, Pigeon2016}.
Similarly, the evolutionary potential of small populations was dismissed by population matrix simulations while ignoring genetic-based arguments (see chapter \ref{chap:dynhet}). Moreover, methods based on phenotypic covariances do not distinguish between the presence and the absence of heritable variation, and cannot be used alone to predict an evolutionary response (chapter \ref{chap:decpop}).

Therefore, all the chapters of this thesis illustrate that a genetic approach, be it based on quantitative genetics or population genetics, is necessary to measure evolution, and can more reliably identify the selective causes and constraints shaping adaptation. 
Attempts to understand the evolutionary dynamics of natural populations based on phenotypic observations only \parencite[e.g.][]{Smallegange2013} are a gamble that might pay off on special occasions, but is unlikely to be reliable in general. 

\subsection{Demographic response to environmental change}
This thesis is almost exclusively concerned with traits and their evolutionary dynamics. 
In the context of understanding the response of natural population to environmental change, such an investigation is legitimate: Whether a trait distribution changes through demographic, plastic, or genetic mechanisms has different consequences on the fate of the population \parencite{Chevin2010}.
Nevertheless, for most applications, and to the eyes of the society, it is unimportant whether animal and plant populations respond to climate change primarily through migration, through plastic changes, or through evolution. Instead, the primary motivation of the research on the response to environmental change is to ascertain whether populations will persist or go extinct, and how managers can affect the outcome.

The question of the persistence of a population is primarily a demographic one. The evolutionary approach that was mine during this PhD is not sufficient to ascertain the fate of the snow vole population, but it might be a useful first step. Indeed, it is now widely acknowledged that evolutionary processes can act on the same time scale as ecological ones, and that they can significantly affect demographics \parencite{Hairston2005,Ellner2011, Chevin2010a, Turcotte2011}. For instance, theory and laboratory experiments support the existence of \emph{evolutionary rescue}, that is, adaptive genetic change within a population that prevent the extinction of the population \parencite{Gonzalez2013a, Schiffers2013a}. Still, empirical evidence for evolutionary rescue in the wild remains extremely limited \parencite{Wal2013}.

In chapter \ref{chap:stasis}, I inferred that the genetic response to selection tended to increase mean juvenile survival over the study period. All other things being equal, evolution therefore had a positive demographic effect and contributed to the recovery of population size. As of yet, however, it is unclear how to quantify the demographic effect of evolution. To the best of my knowledge, an appropriate methodological framework is still lacking. 
Indeed, traditional demographic models used to predict population resilience ignore individual heterogeneity and genetic change \parencite{Kendall2011, vindenes2015, Plard2016}. On the other hand, quantitative genetic studies focus on estimating rates of evolutionary change, but mostly ignore their possible consequences for the dynamics of populations \parencite{Coulson2010,Chevin2012}.
It is now acknowledged that the integration of evolutionary and demographic aspects is crucial for predicting trait dynamics, population resilience and viability \parencite{Schoener2011,Pelletier2012,Chevin2012,Merila2014}. However, only in the last year have methods that could start to address this question in the wild been proposes \parencite{vindenes2015, Coulson2015, Childs2016}, and these should certainly be followed up. 

\section{General conclusion}
Natural selection is a potent force that shapes the evolution of natural populations, but its causes and consequences can be blurred by the complexity of natural populations.
Understanding the process of adaptation requires isolating selection from the stochasticity in fitness components, to disentangle evolution from other drivers of phenotypic change, and to mechanistically link genetic change to selective pressures.
Being able to do so, thanks to an individual-based monitoring including genetic relatedness, I provided a rare example of contemporary adaptive evolution. More examples of evolution in action certainly await to be described, and will make our understanding of the response to environmental change more general and more predictive. This thesis shows in several ways that crucial insight is more likely to come from studies that explicitly study the genetic aspects of selection and phenotypic changes.


%\printbibliography[heading=subbibliography]
